Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-25 paper.Rmd
note
fix re-writing the forecast scoring section

The way civil war is operationalized by Sambanis is so that a country at any given point in time is either at civil war or not, and it cannot experience another civil war onset when it is already at war. In the data, ongoing civil wars are coded as having a missing outcome value, since a new onset is conceptually impossible, and dropped. If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), we would presumable already have observed civil war state for the second half of 2015 and seen that 15 countries had ongoing conflicts. In all previous analyses in the paper, these kinds of cases drop out of the data. However, B&S re-introduce, produce forecasts for them, and include them as positive events ("1"s) in their confusion matrices. They write, on page 21:

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_. Including those cases in 2016-H1 risk set that we are forecasting for would only make sense if all of them had seen civil war termination in 2015-H2, but when making the forecast we would have already known this is not the case.

When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1"). In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g. two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data. Thus it is not possible to treat incidence of civil war similar to onset of civil war, which is what the model actually predicts.

We replicate the B&S confusion matrices but distinguish between non-evens that correspond to countries with no civil war, those with an ongoing civil war, and those with an actual civil war onset in Table \ref{tab:table4}. The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. It is now more apparent that essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset. The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.

With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix, B\&S report that there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives). With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, they report that there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses. Fifteen to 16 civil war _onsets_ in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war _onsets_. Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e. counting ongoing civil wars as if they were onsets.

@@

The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}. In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1, according to the civil conflict dataset they rely on. As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives. This is the norm when predicting rare events, and thus not unusual at all. The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding. This is, at face value, a credit to the model's accuracy.

>>>>

For reference for the following discussion, we reproduce confusion matrices for the 2016-H1 forecasts in Table \ref{tab:table4}, but distinguish between civil war onsets ($Y=1$), non-events ($Y=0$) that correspond to countries with an ongoing civil war, and non-events ($Y=0$) that correspond to countries without a civil war, but which could potentially experience a civil war onset. The original B&S confusion matrices collapse the second two rows in each matrix and treat ongoing and onset as part of the same group. Essentially all of the apparent true positives ($P=1, Y=1$) in the original B&S confusion matrices are actually ongoing civil war cases which by definition _cannot_ experience another civil war onset. The alternative outcome coding B&S evaluate, under "Assuming Change", produces the only two civil war onsets we would have observed in 2016-H1.

@@

The fact that countries with an ongoing civil war cannot experience another civil war onset follows from the way the Sambanis civil war data conceptualize and operationalize civil war. Civil war appears to be a binary property of a country at any given moment in time, i.e. a country at any given point is either at civil war or not, and onset is the first time period corresponding to this state. In the data, ongoing civil wars are coded as having a missing outcome value, and dropped.

If we place ourselves back to the beginning of 2016, when the forecasts were made (the forecasts are pre-registered in early March), this means that we would have observed that there are 15 countries that had ongoing conflicts in the second half of 2015 which had not terminated prior to the forecasting period. In all previous analyses in the paper, these kinds of cases drop out of the data. However, B&S re-introduce, produce forecasts for them, and include them as positive events ("1"s) in their confusion matrices. They write, on page 21:

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_. Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete "win" for the model? When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1"). In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g. two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.

If we turn back to the confusion matrices to re-assess the accuracy of the 2016-H1 forecasts, we can thus conclude that almost all of the predictions are false positives. There are no onsets in the "Assuming Persistence" coding, and two onsets in the "Assuming Change" coding. This is normal. Civil war onsets are infrequent, and it is to be expected that forecasts of rare events will mostly consist of false positives. The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding. This is, at face value, a credit to the model's accuracy.
